---
title: "Metropolis–Hastings MCMC: Sampling the Laplace Distribution"
subtitle: "Random walk Metropolis with Gelman–Rubin convergence diagnostics"
author: "Akrem"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi       = 120
)
```

```{r libraries}
library(ggplot2)
library(scales)   # for axis formatting

# Consistent ggplot2 theme across all plots
theme_set(
  theme_bw(base_size = 11) +
    theme(
      panel.grid.minor  = element_blank(),
      plot.title        = element_text(size = 11),
      plot.subtitle     = element_text(size = 10, colour = "grey40"),
      axis.title        = element_text(size = 12)
    )
)

set.seed(42)
```

---

Markov Chain Monte Carlo lets you draw samples from a distribution when direct sampling is impractical — all you need is a function proportional to its density. This document implements the **random walk Metropolis** algorithm (the simplest member of the Metropolis–Hastings family) from scratch in base R, uses it to sample the Laplace distribution, and then applies the **Gelman–Rubin diagnostic** $\hat{R}$ to answer the question every MCMC practitioner faces: *how do you know your chains have actually converged?*

*This project began as university coursework and has since been revised and extended as a personal portfolio piece. A parallel Python implementation lives in `mcmc_metropolis_hastings.ipynb`.*

# Random Walk Metropolis (N = 10,000, s = 1)

## Target density & sampler

The target distribution is the **Laplace(0, 1)** density:

$$f(x) = \frac{1}{2}e^{-|x|}, \quad x \in \mathbb{R}$$

The proposal at each step is $x^* \sim \mathcal{N}(x_{i-1},\, s^2)$.  
We work in log-scale when computing the acceptance ratio to avoid numerical underflow.

```{r target-and-sampler}
# Log-density of Laplace(0,1): log f(x) = -log(2) - |x|
log_target <- function(x) -log(2) - abs(x)

# True density
target <- function(x) 0.5 * exp(-abs(x))

# Random Walk Metropolis sampler
# Returns a vector of length N+1 (including x0)
random_walk_metropolis <- function(N, s, x0 = 0.0, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  samples    <- numeric(N + 1)
  samples[1] <- x0
  for (i in seq(2, N + 1)) {
    x_star    <- rnorm(1, mean = samples[i - 1], sd = s)
    log_ratio <- log_target(x_star) - log_target(samples[i - 1])
    if (log(runif(1)) < log_ratio) {
      samples[i] <- x_star
    } else {
      samples[i] <- samples[i - 1]
    }
  }
  samples
}

# Gelman-Rubin R-hat; within-chain variance uses the 1/N (population) form
compute_rhat <- function(chains) {
  # chains is a list of numeric vectors (each already excluding x0)
  chain_means <- sapply(chains, mean)
  chain_vars  <- mapply(function(c, m) mean((c - m)^2), chains, chain_means)
  W           <- mean(chain_vars)
  M           <- mean(chain_means)
  B           <- mean((chain_means - M)^2)
  sqrt((B + W) / W)
}
```

## Run sampler & compute Monte Carlo estimates

```{r run-sampler-a}
N_a <- 10000
s_a <- 1.0

chain_a   <- random_walk_metropolis(N = N_a, s = s_a, x0 = 0.0, seed = 42)
samples_a <- chain_a[-1]   # x1 ... xN (drop x0)

mc_mean     <- mean(samples_a)
mc_std      <- sd(samples_a)
accept_rate <- mean(diff(chain_a) != 0)

cat(sprintf("Monte Carlo estimate of the mean               : %.5f\n", mc_mean))
cat(sprintf("Monte Carlo estimate of the standard deviation : %.5f\n", mc_std))
cat(sprintf("Acceptance rate                                : %.4f\n",  accept_rate))
cat(sprintf("\nTheoretical mean (Laplace)                     : 0.00000\n"))
cat(sprintf("Theoretical standard deviation (Laplace)       : %.5f\n",  sqrt(2)))
```

The theoretical Laplace(0,1) distribution has mean **0** and standard deviation $\sqrt{2} \approx 1.4142$.  
The Monte Carlo estimates are very close to these values, which is a good sign that the sampler is working correctly.

## Histogram, KDE & true density

```{r plot-histogram-kde, fig.height=5}
x_grid <- seq(-8, 8, length.out = 500)

# KDE using R's built-in density()
kde_fit <- density(samples_a, bw = "nrd0", n = 512)

plot_title <- sprintf(
  "Random Walk Metropolis — Laplace Target\nN = %s,  s = %.1f,  MC mean ≈ %.4f,  MC std ≈ %.4f,  accept rate ≈ %.2f",
  format(N_a, big.mark = ","), s_a, mc_mean, mc_std, accept_rate
)

ggplot() +
  geom_histogram(
    aes(x = samples_a, y = after_stat(density)),
    bins      = 60,
    fill      = "steelblue",
    alpha     = 0.45,
    colour    = "white",
    linewidth = 0.3
  ) +
  geom_line(
    aes(x = kde_fit$x, y = kde_fit$y, colour = "KDE estimate"),
    linewidth = 1.0
  ) +
  geom_line(
    aes(x = x_grid, y = target(x_grid), colour = "True f(x)"),
    linewidth = 1.0,
    linetype  = "dashed"
  ) +
  scale_colour_manual(
    name   = NULL,
    values = c("KDE estimate" = "darkorange", "True f(x)" = "#DC143C"),
    labels = c("KDE estimate", expression(paste("True ", f(x), " = ", frac(1, 2), e^{-abs(x)})))
  ) +
  coord_cartesian(xlim = c(-8, 8)) +
  labs(
    title  = plot_title,
    x      = expression(italic(x)),
    y      = "Density"
  ) +
  theme(
    legend.position = "top",
    legend.text     = element_text(size = 10)
  )
```

---

# Convergence Diagnostic: $\hat{R}$ (Gelman-Rubin)

The Gelman-Rubin statistic compares within-chain and between-chain variance across multiple chains started at different initial values:

$$\hat{R} = \sqrt{\frac{B + W}{W}}, \quad W = \frac{1}{J}\sum_j V_j, \quad B = \frac{1}{J}\sum_j (M_j - \bar{M})^2$$

If $\hat{R}$ is close to 1 (typically $< 1.05$), chains have mixed well and converged to the same distribution.

## $\hat{R}$ for N = 2,000, s = 0.001, J = 4

```{r rhat-small-s}
N_b            <- 2000
s_b            <- 0.001
J              <- 4
initial_values <- c(-5.0, -1.0, 1.0, 5.0)

chains_b <- lapply(seq_along(initial_values), function(j) {
  chain <- random_walk_metropolis(N = N_b, s = s_b, x0 = initial_values[j], seed = 100 + j)
  chain[-1]   # drop x0
})

rhat_b <- compute_rhat(chains_b)

cat(sprintf("R-hat  (N=%d, s=%.3f, J=%d): %.4f\n", N_b, s_b, J, rhat_b))
cat(sprintf("Converged (R-hat < 1.05)?  %s\n",
            ifelse(rhat_b < 1.05, "Yes", "No — poor mixing detected")))
```

With $s = 0.001$, each proposal step is $\sim 10^{-3}$, so chains started at $\{-5, -1, 1, 5\}$ barely move over 2,000 iterations. Between-chain variance dominates, giving $\hat{R} \gg 1.05$.

## Four-chain trace (N = 2,000, s = 0.001, J = 4)

```{r trace-plot, fig.height=7}
# Build a tidy data frame for ggplot
chain_df <- do.call(rbind, lapply(seq_along(chains_b), function(j) {
  data.frame(
    iteration = seq_along(chains_b[[j]]),
    value     = chains_b[[j]],
    chain     = factor(
      sprintf("x0 = %.1f", initial_values[j]),
      levels = sprintf("x0 = %.1f", initial_values)
    )
  )
}))

ggplot(chain_df, aes(x = iteration, y = value, colour = chain)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  facet_wrap(~ chain, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("steelblue", "darkorange", "seagreen", "#DC143C")) +
  labs(
    title    = sprintf("Four chains — N = %d, s = %.3f, J = %d", N_b, s_b, J),
    subtitle = sprintf("R-hat = %.4f — chains stuck near starting values", rhat_b),
    x        = "Iteration",
    y        = expression(x[i]),
    colour   = NULL
  ) +
  theme(
    legend.position  = "none",
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(size = 9)
  )
```

## $\hat{R}$ vs $s$ — log-spaced grid $[0.001,\, 1]$

Now we sweep across a range of $s$ values to see at what point the chains start mixing well.

```{r rhat-vs-s, fig.height=5}
s_grid      <- exp(seq(log(0.001), log(1), length.out = 60))
rhat_values <- numeric(length(s_grid))

for (idx in seq_along(s_grid)) {
  s_val  <- s_grid[idx]
  chains <- lapply(seq_along(initial_values), function(j) {
    chain <- random_walk_metropolis(
      N    = N_b,
      s    = s_val,
      x0   = initial_values[j],
      seed = 200 + (idx - 1) * J + j
    )
    chain[-1]
  })
  rhat_values[idx] <- compute_rhat(chains)
}

rhat_df <- data.frame(s = s_grid, rhat = rhat_values)

# Highlight the s = 0.001 point
highlight <- data.frame(s = s_b, rhat = rhat_values[1])

ggplot(rhat_df, aes(x = s, y = rhat)) +
  geom_ribbon(
    aes(ymin = 1.0, ymax = pmin(rhat, 1.05)),
    fill = "seagreen", alpha = 0.15
  ) +
  geom_hline(yintercept = 1.05, colour = "#DC143C",
             linetype = "dashed", linewidth = 1.2) +
  geom_line(colour = "steelblue", linewidth = 1.5) +
  geom_point(
    data    = highlight,
    aes(x = s, y = rhat),
    colour  = "darkorange",
    size    = 3.5,
    shape   = 16
  ) +
  annotate(
    "text",
    x     = s_b * 2.5,
    y     = highlight$rhat * 0.97,
    label = sprintf("s = 0.001\nR-hat = %.2f", highlight$rhat),
    hjust = 0,
    size  = 3.2,
    colour = "darkorange"
  ) +
  annotate(
    "text",
    x      = 0.12,
    y      = 1.03,
    label  = "Converged region (R-hat < 1.05)",
    size   = 3.0,
    colour = "seagreen4"
  ) +
  scale_x_log10(
    breaks = c(0.001, 0.01, 0.1, 1),
    labels = c("0.001", "0.01", "0.1", "1")
  ) +
  labs(
    title    = expression(paste("Gelman-Rubin ", hat(R), " Diagnostic vs. Proposal Scale ", italic(s))),
    subtitle = sprintf("N = %d,  J = %d,  initial values = {-5, -1, 1, 5}", N_b, J),
    x        = expression(paste("Proposal standard deviation ", italic(s), " (log scale)")),
    y        = expression(hat(R))
  )
```

---

# Conclusion

The Random Walk Metropolis sampler targets the Laplace(0, 1) distribution well when the proposal scale is chosen appropriately:

- For **$s = 1$**: the acceptance rate is $\approx `r round(accept_rate, 2)`$, the MC mean is $\approx `r round(mc_mean, 4)`$, and the MC standard deviation is $\approx `r round(mc_std, 4)`$ — both very close to the theoretical values of 0 and $\sqrt{2} \approx 1.4142$.

- For **$s = 0.001$**: proposal steps are far too small, so chains barely move from their starting values. The Gelman-Rubin statistic explodes to $\hat{R} \approx `r round(rhat_values[1], 1)`$, far above the convergence threshold of 1.05.

- The $\hat{R}$ vs $s$ sweep shows that convergence is achieved once $s \gtrsim 0.1$. This makes sense because the Laplace(0,1) target has standard deviation $\sqrt{2} \approx 1.41$, so a proposal scale in that ballpark leads to efficient exploration.
```
